---
title: "Multiple Linear Regression"
author: "Osman Omer Mustafa"
date: "today"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    toc-location: left
---

## Load required libraries
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggfortify)
library(car)
library(corrplot)
```

## Read data
```{r message=FALSE}
data <- read_csv("insurance.csv")
```

## Basic checks
```{r}
glimpse(data)
summary(data)
sum(is.na(data))
```

## Ensure categorical variables are factors
```{r}
data <- data %>% 
  mutate(
    sex = as.factor(sex),
    smoker = as.factor(smoker),
    region = as.factor(region)
  )

glimpse(data)
```

Categorical variables (sex, smoker, region) have been converted to factors for proper regression analysis.

## Exploratory Data Analysis

### Summary statistics by smoker status
```{r}
summary <- data %>% 
  group_by(smoker) %>% 
  summarise(
    n = n(),
    mean_charges = mean(charges),
    sd_charges = sd(charges),
    median_charges = median(charges)
  )
summary %>% knitr::kable(caption = "Summary Statistics by Smoker Status")
```

Smokers have substantially higher insurance charges (mean = $32,050) compared to non-smokers (mean = $8,434), 
indicating smoking status is a major predictor of insurance costs.

### Correlation matrix
```{r}
cor_matrix <- data %>% 
  select(age, bmi, children, charges) %>% 
  cor()
cor_matrix %>% knitr::kable(digits = 2, caption = "Correlation matrix")
```

Age shows the strongest correlation with charges (0.30), followed by BMI (0.20). 
Children has minimal correlation (0.07).Correlation analysis provides a preliminary indication of potential multicollinearity, 
which is formally assessed later using Variance Inflation Factor (VIF).

### Correlation plot
```{r}
corrplot(cor_matrix, method = "number", type = "upper")

```

Visual representation confirms age and BMI have moderate positive correlations with charges, 
while children shows negligible association.

### Distribution of charges
```{r}
ggplot(data, aes(x = charges)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(
    x = "Insurance Charges (USD)",
    y = "Frequency",
    title = "Distribution of Insurance Charges"
  )+
  theme(plot.title = element_text(hjust = 0.5))
```

The distribution is right-skewed with a peak at lower charges and a long tail extending to higher values. 
This is typical for insurance data where most individuals have moderate costs while few have very high expenses.

### Charges by smoker status
```{r}
ggplot(data, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  labs(
    x = "Smoker Status",
    y = "Insurance Charges (USD)",
    title = "Insurance Charges by Smoker Status"
  ) +
  theme(legend.position = "none")+
  theme(plot.title = element_text(hjust = 0.5))
```

The boxplot clearly illustrates the substantial difference in charges between smokers and non-smokers. 
Smokers' median charges are approximately four times higher. Some non-smokers appear as outliers with charges 
comparable to smokers, likely due to other risk factors.

### Age vs Charges by smoker
```{r message=FALSE}
ggplot(data, aes(x = age, y = charges, color = smoker)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Age (Years)",
    y = "Insurance Charges (USD)",
    title = "Age vs Charges by Smoker Status",
    color = "Smoker"
  )+
  theme(plot.title = element_text(hjust = 0.5))
```

Both groups show positive relationships between age and charges, but smokers consistently pay higher premiums 
across all age groups. The parallel trend lines suggest age effects are similar for both groups, 
though at different baseline levels.

### BMI vs Charges by smoker
```{r message=FALSE}
ggplot(data, aes(x = bmi, y = charges, color = smoker)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Body Mass Index (BMI)",
    y = "Insurance Charges (USD)",
    title = "BMI vs Charges by Smoker Status",
    color = "Smoker"
  )
```

BMI shows a positive association with charges for both groups. The relationship appears stronger among smokers, 
This pattern suggests a potential interaction between BMI and smoking status, 
which could be formally tested in future models using interaction terms.

## Models

### Simple Linear Regression
Fit a baseline model using age as the sole predictor.
```{r}
model_simple <- lm(charges ~ age, data = data)
tidy_res <- tidy(model_simple)
tidy_res %>% knitr::kable(caption = "Simple Linear Regression")
```

### Multiple Linear Regression
Fit a full model including all available predictors.
```{r}
model_full <- lm(charges ~ age + bmi + sex + children + smoker + region, data = data)
tidy_res2 <- tidy(model_full)
tidy_res2 %>% knitr::kable(caption = "Multiple Linear Regression")
```

### Model performance metrics
```{r}
glance(model_simple) %>% knitr::kable(caption = "Simple Model Performance")
glance(model_full) %>% knitr::kable(caption = "Multiple Model Performance")
```

The multiple regression model shows substantial improvement over the simple model, with R² increasing from 0.089 to approximately 0.75, 
indicating the full model explains 75% of variability in insurance charges.

### Model comparison
```{r}
anova_res <- anova(model_simple, model_full)
anova_res %>% knitr::kable(caption = "Anova Result")
AIC_res <- AIC(model_simple, model_full)
AIC_res %>% knitr::kable(caption = "AIC Result")
BIC_res <- BIC(model_simple, model_full)
BIC_res %>% knitr::kable(caption = "BIC Result")
```

ANOVA confirms the multiple model is significantly better than the simple model (p < 0.001). 
Both AIC and BIC are substantially lower for the full model, further supporting its superiority.

## Diagnostic plots

### Residuals vs Fitted
Check for linearity and homoscedasticity.
```{r warning=FALSE , message=FALSE}
autoplot(model_full, which = 1) +
  labs(title = "Residuals vs Fitted (Multiple Regression)")
```

Residuals show relatively random scatter around zero, though some funneling is visible at higher fitted values, 
suggesting mild heteroscedasticity. Overall, the linearity assumption appears reasonably satisfied for most observations, 
with minor deviations at higher fitted values.


### Normal Q-Q
Assess normality of residuals.
```{r}
autoplot(model_full, which = 2) +
  labs(title = "Normal Q-Q Plot")
```

Most points follow the diagonal line closely, indicating approximate normality. Some deviation appears at the tails, 
particularly for extreme values, which is common in real-world data.

### Scale-Location
Examine homoscedasticity assumption.
```{r}
autoplot(model_full, which = 3) +
  labs(title = "Scale-Location Plot")
```

The scale-location plot shows slight upward trend, confirming mild heteroscedasticity. 
The variance of residuals increases somewhat with fitted values, though not severely.

### Cook's Distance
Identify influential observations.
```{r}
autoplot(model_full, which = 5) +
  labs(title = "Cook's Distance")
```

No single observation appears to exert undue influence on the overall model estimates.
, indicating minimal influence of individual points on model coefficients. 
The model is not unduly affected by outliers.

## Multicollinearity check
```{r}
vif_values <- vif(model_full)
vif_values %>% knitr::kable(caption = "Multicollinearity check")
```

All VIF values are well below 5, indicating no multicollinearity concerns among predictors. 
Each variable contributes unique information to the model.

## Coefficients table
```{r}
coef_table <- tidy(model_full, conf.int = TRUE)
coef_table %>% 
  mutate(across(where(is.numeric), ~round(., 3))) %>%
  knitr::kable(caption = "Regression Coefficients with 95% Confidence Intervals")
```

## Conclusion

The multiple linear regression analysis reveals several key findings:

**Model Performance:** The full model achieves R² = 0.75, explaining three-quarters of variability in insurance charges. 
This represents substantial improvement over the simple age-only model (R² = 0.09).

**Key Predictors:** Smoking status emerges as the dominant predictor, with smokers paying approximately $23,800 more than non-smokers, 
holding all other factors constant. Age and BMI also show significant positive effects, with each additional year increasing charges 
by roughly $257, and each BMI unit adding approximately $340.

**Model Assumptions:** Diagnostic plots indicate the model satisfies key regression assumptions reasonably well. 
Mild heteroscedasticity is present but not severe. Residuals approximate normality with slight tail deviations. 
No multicollinearity issues exist among predictors.

**Practical Implications:** The model demonstrates that smoking status, age, and BMI are critical factors in determining insurance costs. 
These findings align with actuarial expectations and provide a solid foundation for understanding premium determinants.

**Limitations:** The right-skewed nature of charges and presence of outliers suggest some individuals have unusually high costs 
that may warrant separate investigation. Future analyses could explore interaction effects, particularly between smoking and other risk factors.

Although the model explains a large proportion of variance, causal inference is not possible 
due to the observational nature of the data.
